Real Time Variational Density Forecasting

### 2017-02-27 Confirmation/confirmation_V2.tex
intro changes

%2multibyte Version: 5.50.0.2960 CodePage: 1252
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[centertags]{amsmath}
\usepackage{graphicx}%
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[hidelinks]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{amsmath}
\usepackage[]{algorithm2e}
\usepackage{amsthm}
\usepackage{url}
\usepackage{wasysym}
\usepackage{ulem}
\usepackage{tikz}
%\usetikzlibrary{bayesnet}
\usepackage{afterpage}
\usepackage{bbm}
\setcounter{MaxMatrixCols}{30}
\providecommand{\U}[1]{\protect\rule{.1in}{.1in}}
\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0.1in}
\setlength{\evensidemargin}{0.1in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{8.25in}
\numberwithin{equation}{section}
\title{Real Time Variational Density Forecasting}
\author{Nathaniel Tomasetti, Catherine Forbes and Anastasios Panagiotelis}
\begin{document}

@@

Bayesian methods are computationally complex as they require integration over the set of unknowns, whether parameters or latent stochastic variables. This set is quite large for models used to model short term electricity load and price as there are strong seasonal patterns in the data \citep{Taylor2003}. While analytical integration is typically not feasible to obtain the predictive distribution for such models, however there are a range of numerical and simulation-based techniques available for this purpose, notably Markov Chain Monte Carlo (MCMC) methods. While it is often possible to demonstrate theoretical convergence of MCMC algorithms, computation may be slow for even moderately complex models. This is problematic for when attempting to forecast electricity load in Victoria, when realizations occur at five minute intervals, the predictive density for the next period's electricity load may not be available before it is observed.

@@

The discussion of MCMC and VB will be continued in Section \ref{sec:BayesInf}, with the methods contrasted in the context of two simple models. Section \ref{sec:ElecLoad} will introduce a planned emperical application of VB to Victorian electricity load using an exponential smoothing model and data from the 2012-2015 period. Section \ref{sec:Discussion} outlines the major areas to which this thesis will make a contribution, and Section \ref{sec:Timeline} outlines the planned timeline for these contributions.

@@

To facilitate the discussion, consideration is given to models where the unknowns may be summarized in a (finite) $k-$dimensional vector, denoted by $\boldsymbol{\theta}.$ In this setting, given an observed time series, denoted by $y_{1:T}= \{y_{t}, t = 1, \dots T\}$, the Bayesian forecast distribution associated with time $t=T+1$ is characterized by the conditional density
\begin{equation}
\label{predictive}
p(y_{T+1} | y_{1:T}) =\int p(y_{T+1}|y_{1:T},\boldsymbol{\theta}) p(\boldsymbol{\theta}|y_{1:T}) d\boldsymbol{\theta},
\end{equation}
where $p(\boldsymbol{\theta}|y_{1:T})$ denotes the posterior density for $\boldsymbol{\theta}$, given by
\begin{equation}
\label{posterior}
 p(\boldsymbol{\theta} | y_{1:T}) = \frac{p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta})}{\int_{\boldsymbol{\theta}}p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta}) d\boldsymbol{\theta}}.
\end{equation}

Generally the analytical solution to (\ref{posterior}), and hence to (\ref{predictive}) will be unavailable due to the complex functional form of the distributions involved. 
In this section two alternative methods for computing the desired posterior distribution in (\ref{posterior}),  MCMC and VB, will be reviewed.
In brief, MCMC is used to create a sample from $p(\boldsymbol{\theta} | y_{1:T})$, with any function of $\boldsymbol{\theta}$ that is desired estimated from that sample. In contrast, VB replaces $p(\boldsymbol{\theta} | y_{1:T})$ with an approximation, denoted by $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, where $\boldsymbol{\lambda}$ is a vector of auxiliary parameters associated with the approximation and may depend on the observations $y_{1:T}$.
For example, if $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is Gaussian, then $\boldsymbol{\lambda}$ would denote the mean and variance parameters that characeterise this normal distribution for $\boldsymbol{\theta}$.  Variational Bayes aims to choose the distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ such that the Kullback-Leibler divergence from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to $p(\boldsymbol{\theta} | y_{1:T})$ is minimised. Once the posterior distribution in (\ref{posterior}) is available, the predictive density forecast (\ref{predictive}) can be found by simulation.

@@

There are many types of MCMC algorithms, with arguably the simplest and most commonly used one being the Gibbs sampler. The Gibbs sampler algorithm iteratively samples the components of the $k-$dimensional parameter vector $\boldsymbol{\theta}$ via each of the so-called full conditional distributions as follows,
\begin{align}
&p(\theta_1 | \theta_2, \dots, \theta_k, y_{1:T}) \nonumber \\
&p(\theta_2 | \theta_1, \theta_3, \dots, \theta_k, y_{1:T}) \nonumber \\
&\vdots \nonumber \\
&p(\theta_k | \theta_1, \dots, \theta_{k-1}, y_{1:T}). \nonumber
\end{align}
Under mild regularity conditions (see, e.g., \citet{Tierney1994}) and with enough iterations of the Markov chain that results from the Gibbs sampler, these samples converge in distribution to $p(\boldsymbol{\theta} | y_{1:T})$. Samples taken before the MCMC converges ot the posterior must be discarded, and the remaining samples may have strong dependence between consecutive draws of the same parameter due to the Markov nature of the algorithm. The computation time for each iteration and the overall number of iterations required to accruately summarise the posterior distribution is problem specific and typically increases with the number of parameters in the model. On occasion, the full conditional distribution can not be recognised and a Metropolis-Hastings-within-Gibbs step can be used, see \citet{Gilks1995} for details.

For illustration, consider data $y_t$ for $t = 1, \dots, T$ generated independently from a $\mathcal{N}(\mu, \sigma^2)$ distribution with prior distributions given by $\mu | \sigma^2 \sim \mathcal{N}(\gamma, \sigma^2/\tau)$ and $\sigma^2 \sim Inv.Gamma(\mbox{shape} = \alpha, \mbox{scale =}\beta)$. It can be shown that

\begin{align}
&\mu | \sigma^2, y_{1:T} \sim \mathcal{N} \left(\frac{\tau\gamma + T\bar{y}}{\tau + T}, \frac{\sigma^2}{\tau  + T} \right) \label{posmu} \\
&\sigma^2 | y_{1:T} \sim Inv.Gamma  \left( \mbox{shape} = \alpha + T/2, \mbox{scale} = \beta + \frac{1}{2}\sum_{t=1}^{T}y_t^2 + \tau\gamma^2 - \frac{(\tau\gamma + T\bar{y})^2}{\tau + T} \right) \label{possig2} 
\end{align}
where $\bar{y} = 1/T \sum_{t=1}^{T}y_t$. Posterior sampling can proceed by first simulating $\sigma^2$ from (\ref{possig2}) and conditioning on this draw to simulate $\mu$ from (\ref{posmu}). It should be noted that as we are simulating from $p(\sigma^2 | y)$ rather than $p(\sigma^2 | \mu, y)$, there is no Markov dependency between iterations and this is a Monte Carlo sampler rather than an MCMC.

@@

As discussed in Section \ref{sec:Intro}, a faster (albeit approximate) alternative to MCMC is VB. This approach introduces an approximating distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ with the choice of the functional form $q$ and subsequent auxiliary parameter vector $\boldsymbol{\lambda}$ determined so that the Kullback-Leibler (KL) divergence \citep{Kullback1951} from$q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ the true posterior $p(\boldsymbol{\theta} | y_{1:T})$ is minimised. The KL divergence is defined by
\begin{equation}
\label{KL-def}
KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T}] = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta} |y_{1:T})}\right) d\boldsymbol{\theta}.
\end{equation}
The KL divergence is a non-negative, asymetric measure of the discrepancy between $p(\boldsymbol{\theta} | y_{1:T})$ and $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$  that will be equal to zero if and only if $p(\boldsymbol{\theta} | y_{1:T}) = q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ almost everywhere \citep{Bishop2006}. Note that $KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})]$ can be expressed as
\begin{equation}
\label{KL-ELBO}
KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q, y_{1:T})
\end{equation}
where $\mathcal{L}(q, y_{1:T})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T}))$.  $\mathcal{L}(q, y_{1:T})$ is defined by
\begin{equation}
\label{ELBO}
\mathcal{L}(q, y_{1:T}) = \int_{\boldsymbol{\theta}} q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (p(y_{1:T},\boldsymbol{\theta})) d\boldsymbol{\theta} -  \int_{\boldsymbol{\theta}} q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (q(\boldsymbol{\theta} | \boldsymbol{\lambda})) d\boldsymbol{\theta}.
\end{equation}

Since $\ln(p(y_{1:T}))$ is constant with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, maximising (\ref{ELBO}) with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is equivalent to minimising (\ref{KL-def}). Maximising the ELBO is much more convenient than minimising the KL Divergence, as will be shown in the following sections in the context of the two major implementations of Varitaional Bayes, namely Mean Field Variational Bayes and Stochastic Variational Bayes. Each of these implementations take advantage of the functional form of the ELBO, and offer computationally efficient algorithms to find a $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that is optimal within a particular class of distributions.

@@

Mean Field Variational Bayes (MFVB) has origins in physics \citep{Chandler1987} and has since developed a large literature in machine learning, notably developed in \citet{Jordan1999} and \citet{Beal2003}. MFVB restricts the class of distributions from which to select $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to the set of factorisable distributions,
\begin{equation}
\label{mf1}
q(\boldsymbol{\theta} | \boldsymbol{\lambda}) = \prod_{i=1}^k q_i(\theta_i | \boldsymbol{\lambda_i}).
\end{equation}

Each of the $k$ components $\theta_i$ may be a vector, but in most implementations are scalars. Each $\theta_i$ has an associated vector $\boldsymbol{\lambda_i}$ which may be of a different dimension to $\theta_i$. $\boldsymbol{\lambda_i}$ is an auxillary parameter vector for the relevant factor $q_i$, which will be used in this section as shorthand notation for the distribution $q_i(\theta_i |\boldsymbol{\lambda_i})$.

MFVB is widely used as it greatly simplifies maximisation of the ELBO, and has a straight-forward implementation in exponential family models \citep{Wainwright2008}.  Maximising the ELBO with respect to $q_i$ is as analytically involved as deriving the conditional distributions used in Gibbs based MCMC schemes, but the resulting algorithm to maximise the ELBO is much simpler computationally than sampling from a Gibbs MCMC scheme. MFVB factorises (\ref{ELBO}) into a product of $k$ seperate components that each contain only one $q_i$, and then maximises each of these factors with respect to the associated $q_i$. Using the notation $q_{\setminus i} = \prod_{j\neq i}q_j$, \citet{Attias1999} shows that each component is maximised by
\begin{equation}
\label{mf2}
q_i(\theta_i |\boldsymbol{\lambda_i}) \propto\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y_{1:T},\boldsymbol{\theta}))]).
\end{equation}

We illustrate the implementation of MFVB with the model considered in Section (\ref{sec:MCMC}), where $y_t$ for $t = 1, \dots, T$ is generated independently from a $\mathcal{N}(\mu, \sigma^2)$ distribution, and prior distributions are given by $\mu \sim \mathcal{N}(\gamma, \sigma^2 / \tau)$ and $\sigma^2 \sim Inv.Gamma(\mbox{shape} = \alpha, \mbox{scale =}\beta)$. It can be shown that
\begin{equation}
\label{mf3}
q(\mu | \boldsymbol{\lambda}_1) \propto \exp \left\{ \frac{-(T + \tau) \mathbb{E}_{q(\sigma^2)}(\sigma^{-2})}{2} \left( \mu - \frac{\tau \gamma + \sum_{t=1}^{T} y_t}{T + \tau} \right)^2 \right\}
\end{equation}
and
\begin{align}
\label{mf4}
q(\sigma^2 | \boldsymbol{\lambda}_2) &\propto \sigma^{-2((T+1)/2 + \alpha + 1)} \nonumber \\
\times &\exp \left\{ \frac{ \beta + 1/2\left((\tau + T)\mathbb{E}_{q(\mu)} - 2 \mathbb{E}_{q(\mu)}(\mu)(\sum_{t=1}^{T}y_t + \tau) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2\right)}{-\sigma^2} \right\}.
\end{align}

As the prior distributions used were conjugate to the likelihood, the Variational Bayes optimal distributions for $q(\mu |  \boldsymbol{\lambda}_1)$ and $q(\sigma^2 | \boldsymbol{\lambda}_2)$ are the same family as the prior distributions, and are denoted by $q(\mu | \boldsymbol{\lambda}_1) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\tau})$ and $q(\sigma^2 | \boldsymbol{\lambda}_2) \sim Inv.Gamma(\tilde{\alpha}, \tilde{\beta})$. The vector $\boldsymbol{\lambda} = (\tilde{\gamma}, \tilde{\tau}, \tilde{\alpha}, \tilde{\beta})'$ is given by
\begin{align}
\tilde{\gamma} &= \frac{\tau \gamma + \sum_{t=1}^{T} y_t}{T + \tau}  \label{mf5} \\ 
\tilde{\tau} &= \left((T + \tau)\mathbb{E}_{q(\sigma^2)}(\sigma^{-2})\right )^{-1} \label{mf6} \\
\tilde{\alpha} &= (T+1)/2 + \alpha  \label{mf7} \\
\tilde{\beta} &= \beta + 1/2\left((\tau + T)\mathbb{E}_{q(\mu)} - 2 \mathbb{E}_{q(\mu)}(\mu)(\sum_{t=1}^{T}y_t + \tau) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right). \label{mf8}
\end{align}.

>>>>

%2multibyte Version: 5.50.0.2960 CodePage: 1252
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[centertags]{amsmath}
\usepackage{graphicx}%
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[hidelinks]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{amsmath}
\usepackage[]{algorithm2e}
\usepackage{amsthm}
\usepackage{url}
\usepackage{wasysym}
\usepackage{ulem}
\usepackage{afterpage}
\usepackage{bbm}
\setcounter{MaxMatrixCols}{30}
\providecommand{\U}[1]{\protect\rule{.1in}{.1in}}
\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0.1in}
\setlength{\evensidemargin}{0.1in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{8.25in}
\numberwithin{equation}{section}
\title{Real Time Variational Density Forecasting}
\author{Nathaniel Tomasetti}
\begin{document}

@@

Bayesian methods are computationally complex as they require integration over the set of unknowns, whether parameters or latent stochastic variables. This set is typically quite large for models used to model short term electricity load and price as there are strong seasonal patterns in the data \citep{Taylor2003}. Analytical integration is typically not feasible to obtain the predictive distribution for such models, however there are a range of numerical and simulation-based techniques available for this purpose, notably Markov Chain Monte Carlo (MCMC) methods. While it is often possible to demonstrate theoretical convergence of MCMC algorithms, computation may be slow for even moderately complex models. This is problematic when attempting to forecast electricity load in Victoria, as realizations occur at five minute intervals, the predictive density for the next period's electricity load may not be available before it is observed.

@@

The literature on Variational Bayes is dominated by two major implementations: Mean Field Variational Bayes (MFVB) and what we call Stochastic Variational Bayes (SVB). For either implementation we emphasise that there are two seperate tasks: Choose a functional form for the approximating distribution, and then optimise the parameters of this distribution to minimise a well defined divergence between the approximation and the true posterior.

Mean Field Variational Bayes (see \citet{Bishop2006} for an overview) can be applied to exponetial family models with conjugate priors, and restricts the class of approximating distributions to the factorisable family. MFVB provides both the optimal form of the approximating distribution within this class, and a computationally simple coordinate ascent algorithm is available to optimise parameters.
Stochastic Variational Bayes (see \citet{Paisley2012} and \citet{Ranganath2014}) lifts both the restrictions for an exponential family model and the use of an approximating distribution that is factorisable. While parameter optimisation has a well developed gradient ascent algorithm, the determination of the functional form to be used is an active area of research.

We explore the use of a copula to augment the approximating distribution with a flexible dependence structure, which has seen success in \citet{Tran2015} and \citet{Ng2016}. The time series context of our implementation implies that MCMC samples for out of date samples are available, and we emphasise the use of this information to guide the construction of a vine copula so that the approximation has a dependence structure similiar to the true posterior.

The discussion of MCMC and VB will be continued in Section \ref{sec:BayesInf}, with the methods contrasted in the context of two simple models. Section \ref{sec:ElecLoad} will introduce a planned empirical application of VB to Victorian electricity load using an exponential smoothing model and data from the 2012-2015 period. Section \ref{sec:Discussion} outlines the major areas to which this thesis will make a contribution, and Section \ref{sec:Timeline} outlines the planned timeline for these contributions.

@@

To facilitate the discussion, consideration is given to models where the unknowns may be summarized in a (finite) $k-$dimensional vector, denoted by $\boldsymbol{\theta}.$ In this setting, given an observed time series, denoted by $y_{1:T}= \{y_{t}, t = 1, \dots T\}$, the Bayesian forecast distribution associated with time $t=T+1$ is characterized by the conditional density
\begin{equation}
\label{predictive}
p(y_{T+1} | y_{1:T}) =\int p(y_{T+1}|y_{1:T},\boldsymbol{\theta}) p(\boldsymbol{\theta}|y_{1:T}) d\boldsymbol{\theta},
\end{equation}
where $p(\boldsymbol{\theta}|y_{1:T})$ denotes the posterior density for $\boldsymbol{\theta}$, given by
\begin{equation}
\label{posterior}
 p(\boldsymbol{\theta} | y_{1:T}) = \frac{p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta})}{\int p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta}) d\boldsymbol{\theta}}.
\end{equation}

Generally the analytical solution to (\ref{posterior}), and hence to (\ref{predictive}) will be unavailable due to the complex functional form of the distributions involved. 
In this section two alternative methods for computing the desired posterior distribution in (\ref{posterior}),  MCMC and VB, will be reviewed.
In brief, MCMC is used to create a sample from $p(\boldsymbol{\theta} | y_{1:T})$, with any function of $\boldsymbol{\theta}$ that is desired estimated from that sample. In contrast, VB replaces $p(\boldsymbol{\theta} | y_{1:T})$ with an approximation, denoted by $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, where $\boldsymbol{\lambda}$ is a vector of auxiliary parameters associated with the approximation that may depend on the observations $y_{1:T}$.
For example, if $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is Gaussian, then $\boldsymbol{\lambda}$ would denote the mean and variance parameters that characterise this distribution for $\boldsymbol{\theta}$.  Variational Bayes aims to choose the distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ within a restricted class such that the a measure of divergence, typically the Kullback-Leibler divergence from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to $p(\boldsymbol{\theta} | y_{1:T})$, is minimised. Once the posterior distribution in (\ref{posterior}) is available, the predictive density forecast (\ref{predictive}) can be found by simulation.

@@

There are many types of MCMC algorithms, with arguably the simplest and most commonly used one being the Gibbs sampler. The Gibbs sampler algorithm iteratively samples the components of the $k-$dimensional parameter vector $\boldsymbol{\theta}$ via each of the so-called full conditional distributions as follows,
\begin{align}
&p(\theta_1 | \theta_2, \dots, \theta_k, y_{1:T}) \nonumber \\
&p(\theta_2 | \theta_1, \theta_3, \dots, \theta_k, y_{1:T}) \nonumber \\
&\vdots \nonumber \\
&p(\theta_k | \theta_1, \dots, \theta_{k-1}, y_{1:T}). \nonumber
\end{align}
Under mild regularity conditions (see, e.g., \citet{Tierney1994}) and with enough iterations of the Markov chain that results from the Gibbs sampler, these samples converge in distribution to $p(\boldsymbol{\theta} | y_{1:T})$. Samples taken before the MCMC converges to the posterior must be discarded, and the remaining samples may have strong dependence between consecutive draws of the same parameter due to the Markov nature of the algorithm. The computation time for each iteration and the overall number of iterations required to accruately summarise the posterior distribution is problem specific and typically increases with the number of parameters in the model. On occasion, the full conditional distribution cannot be recognised and a Metropolis-Hastings-within-Gibbs step can be used, see \citet{Gilks1995} for details.

For illustration, consider data $y_t$ for $t = 1, \dots, T$ generated independently from a $\mathcal{N}(\mu, \sigma^2)$ distribution with prior distributions given by $\mu | \sigma^2 \sim \mathcal{N}(\gamma, \sigma^2/\tau)$ and $\sigma^2 \sim IG(\mbox{shape} = \alpha, \mbox{scale =}\beta)$, where $IG$ repesents the inverse of a Gamma distribution. It can be shown that

\begin{align}
&\mu | \sigma^2, y_{1:T} \sim \mathcal{N} \left(\frac{\tau\gamma + T\bar{y}}{\tau + T}, \frac{\sigma^2}{\tau  + T} \right) \label{posmu} \\
&\sigma^2 | \mu, y_{1:T} \sim IG  \left( \mbox{shape} = \alpha + T/2, \mbox{scale} = \beta + \frac{1}{2}\sum_{t=1}^{T}y_t^2 + \tau\gamma^2 - \frac{(\tau\gamma + T\bar{y})^2}{\tau + T} \right) \label{possig2} 
\end{align}
where $\bar{y} = 1/T \sum_{t=1}^{T}y_t$. Posterior sampling can proceed by first simulating $\sigma^2$ from (\ref{possig2}) and conditioning on this draw to simulate $\mu$ from (\ref{posmu}).

@@

As discussed in Section \ref{sec:Intro}, a faster (albeit approximate) alternative to MCMC is VB. VB is an umbrella term for a broad collection of algorithms that introduce an approximating distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ with the choice of the functional form $q$ and subsequent auxiliary parameter vector $\boldsymbol{\lambda}$ determined so that an error function is minimised. The classical choice for this error function is the Kullback-Leibler (KL) divergence \citep{Kullback1951} from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ the true posterior $p(\boldsymbol{\theta} | y_{1:T})$. The KL divergence is defined by
\begin{equation}
\label{KL-def}
KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta} |y_{1:T})}\right) d\boldsymbol{\theta}.
\end{equation}
The KL divergence is a non-negative, asymetric measure of the discrepancy between $p(\boldsymbol{\theta} | y_{1:T})$ and $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$  that will be equal to zero if and only if $p(\boldsymbol{\theta} | y_{1:T}) = q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ almost everywhere \citep{Bishop2006}. Note that $KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})]$ can be expressed as
\begin{equation}
\label{KL-ELBO}
KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})
\end{equation}
where $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T}))$.  $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ is defined by
\begin{equation}
\label{ELBO}
\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}) = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (p(y_{1:T},\boldsymbol{\theta})) d\boldsymbol{\theta} -  \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (q(\boldsymbol{\theta} | \boldsymbol{\lambda})) d\boldsymbol{\theta}.
\end{equation}

Since $\ln(p(y_{1:T}))$ is constant with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, maximising (\ref{ELBO}) with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is equivalent to minimising (\ref{KL-def}). Maximising the ELBO is much more convenient than minimising the KL Divergence, as will be shown in the following sections in the context of the two major implementations of Variational Bayes we consider, Mean Field Variational Bayes and Stochastic Variational Bayes. Each of these implementations take advantage of the functional form of the ELBO, and offer computationally efficient algorithms to find a $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that is optimal within a particular class of distributions. The statistical properties of $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ in MFVB and SVB are largely driven by the choice of the KL divergence from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to $p(\boldsymbol{\theta} | y_{1:T})$ as an error function, and there is significant interest in other error functions, such as the KL divergence from $p(\boldsymbol{\theta} | y_{1:T})$ to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ \citep{Minka2001} and Stein's discrepency (\citealp{Liu2016}; \citealp{Ranganath2016b}).

@@

Mean Field Variational Bayes (MFVB) has origins in physics (\citealp{Amari1985}; \citealp{Chandler1987}) and has since developed a large literature in machine learning, notably developed in \citet{Jordan1999} and \citet{Ghahramani2000}. MFVB restricts the class of distributions from which to select $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to the set of factorisable distributions,
\begin{equation}
\label{mf1}
q(\boldsymbol{\theta} | \boldsymbol{\lambda}) = \prod_{i=1}^k q_i(\theta_i | \boldsymbol{\lambda_i}).
\end{equation}

Each of the $k$ components $\theta_i$ may be a vector, but in most implementations are scalars. Each $\theta_i$ has an associated vector $\boldsymbol{\lambda_i}$ which may have different dimension to $\theta_i$. $\boldsymbol{\lambda_i}$ is an auxillary parameter vector for the relevant factor $q_i$, which will be used in this section as shorthand notation for the distribution $q_i(\theta_i |\boldsymbol{\lambda_i})$.

MFVB is widely used as it greatly simplifies maximisation of the ELBO, and has a straight-forward implementation in exponential family models \citep{Wainwright2008}.  Maximising the ELBO with respect to $q_i$ is as analytically involved as deriving the conditional distributions used in Gibbs based MCMC schemes, but the resulting algorithm to maximise the ELBO is much simpler computationally than sampling from a Gibbs MCMC scheme. MFVB factorises (\ref{ELBO}) into a product of $k$ separate components that each contain only one $q_i$, and then maximises each of these factors with respect to the associated $q_i$. Using the notation $q_{\setminus i} = \prod_{j\neq i}q_j$, \citet{Attias1999} shows that each component is maximised by
\begin{equation}
\label{mf2}
q_i(\theta_i |\boldsymbol{\lambda_i}) \propto\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y_{1:T},\boldsymbol{\theta}))]).
\end{equation}

We illustrate the implementation of MFVB with the model considered in Section (\ref{sec:MCMC}), where $y_t$ for $t = 1, \dots, T$ is generated independently from a $\mathcal{N}(\mu, \sigma^2)$ distribution, and prior distributions are given by $\mu \sim \mathcal{N}(\gamma, \sigma^2 / \tau)$ and $\sigma^2 \sim IG(\mbox{shape} = \alpha, \mbox{scale =}\beta)$. It can be shown that
\begin{equation}
\label{mf3}
q(\mu | \boldsymbol{\lambda}_1) \propto \exp \left\{ \frac{-(T + \tau) \mathbb{E}_{q(\sigma^2)}(\sigma^{-2})}{2} \left( \mu - \frac{\tau \gamma + \sum_{t=1}^{T} y_t}{T + \tau} \right)^2 \right\}
\end{equation}
and
\begin{align}
\label{mf4}
q(\sigma^2 | \boldsymbol{\lambda}_2) &\propto \sigma^{-2((T+1)/2 + \alpha + 1)} \nonumber \\
\times &\exp \left\{ \frac{ \beta + 1/2\left((\tau + T)\mathbb{E}_{q(\mu)}(\mu^2) - 2 \mathbb{E}_{q(\mu)}(\mu)(\sum_{t=1}^{T}y_t + \tau) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2\right)}{-\sigma^2} \right\}.
\end{align}

As the prior distributions used were conjugate to the likelihood, the Variational Bayes optimal distributions for $q(\mu |  \boldsymbol{\lambda}_1)$ and $q(\sigma^2 | \boldsymbol{\lambda}_2)$ are the same family as the prior distributions, and are denoted by $q(\mu | \boldsymbol{\lambda}_1) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\tau})$ and $q(\sigma^2 | \boldsymbol{\lambda}_2) \sim IG(\tilde{\alpha}, \tilde{\beta})$. The vector $\boldsymbol{\lambda} = (\tilde{\gamma}, \tilde{\tau}, \tilde{\alpha}, \tilde{\beta})'$ is given by
\begin{align}
\tilde{\gamma} &= \frac{\tau \gamma + \sum_{t=1}^{T} y_t}{T + \tau}  \label{mf5} \\ 
\tilde{\tau} &= \left((T + \tau)\mathbb{E}_{q(\sigma^2)}(\sigma^{-2})\right )^{-1} \label{mf6} \\
\tilde{\alpha} &= (T+1)/2 + \alpha  \label{mf7} \\
\tilde{\beta} &= \beta + 1/2\left((\tau + T)\mathbb{E}_{q(\mu)} - 2 \mathbb{E}_{q(\mu)}(\mu)(\sum_{t=1}^{T}y_t + \tau) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right). \label{mf8}
\end{align}.
